dispRity ecology demo

### 2016-05-29 doc/dispRity-ecology-demo.Rmd

This demo aims to give quick overview of the <span>`dispRity`
</span>package (v.0.2) for ecological analysis. Please refer to [GitHub
page:
github.com/TGuillerme/dispRity](https://github.com/TGuillerme/dispRity)
for other vignettes, namely the `dispRity` tutorial that
explains the functions in more detail.

@@

` `\
`   `\

@@

\

\
\
\
\
\
\
\
\
`  `\
\
\
\
`  `\
`  `

@@

\
\
`   `\
`   `\
\
`   `\
`   `\
\
`   `\
`   `\
\
\
`  `\
\
`          `\
`         `

@@

\
`       `\
`            `\
`   `\

    ##     Treat Depth
    ## 1       a     1
    ## 1.1     a     2
    ## 2       b     1
    ## 2.1     b     2
    ## 3       a     1
    ## 3.1     a     1

@@

\
`  `\
\
\

    ## [1] "dispRity"

\
\

    ## 4 custom series for 40 elements 
    ## Series:
    ## Treat.a, Treat.b, Depth.1, Depth.2.

@@

\
\
\
\
`   `

@@

\
`     `\
\
\
\
\

    ## Disparity measurements across 4 series for 40 elements 
    ## Series:
    ## Treat.a, Treat.b, Depth.1, Depth.2.
    ## Disparity calculated as: c(median, centroids) for 20 dimensions.
    ## Data was split using custom method.

    ##    series  n observed
    ## 1 Treat.a 21    82.38
    ## 2 Treat.b 19    94.68
    ## 3 Depth.1 23    83.98
    ## 4 Depth.2 17    89.72

@@

\
`     `\
\
\
`     `\
`      `\
\
\
\

    ## Bootstrapped ordinated matrix with 40 elements 
    ## Series:
    ## Treat.a, Treat.b, Depth.1, Depth.2.
    ## Data was split using custom method.
    ## Data was bootstrapped 100 times, using the full bootstrap method.

    ## Bootstrapped ordinated matrix with 40 elements 
    ## Series:
    ## Treat.a, Treat.b, Depth.1, Depth.2.
    ## Data was split using custom method.
    ## Data was bootstrapped 100 times, using the full bootstrap method.
    ## Data was rarefied with a maximum of 17 elements

@@

\
`     `\
\
\
`     `

@@

\

    ##    series  n observed  mean  2.5%   25%   75% 97.5%
    ## 1 Treat.a 21    82.38  79.3 64.83 74.45  84.1  95.4
    ## 2 Treat.b 19    94.68 100.2 79.63 90.99 106.8 133.6
    ## 3 Depth.1 23    83.98  82.4 69.04 78.32  85.9  92.7
    ## 4 Depth.2 17    89.72 100.9 69.58 85.03 110.6 163.4

\
\
`     `

    ##    series  n observed median    5%   95%
    ## 1 Treat.a 21    82.38  78.50 68.34  91.2
    ## 2 Treat.b 19    94.68  97.02 83.17 128.3
    ## 3 Depth.1 23    83.98  83.21 72.79  91.3
    ## 4 Depth.2 17    89.72  93.04 75.77 149.4

\

    ##    series  n observed  mean  2.5%   25%   75% 97.5%
    ## 1 Treat.a 17       NA  79.4 67.47 74.52  83.1  96.7
    ## 2 Treat.b 17       NA  99.8 79.24 88.04 106.9 147.4
    ## 3 Depth.1 17       NA  82.4 69.41 77.08  86.3 106.1
    ## 4 Depth.2 17    89.72 102.7 77.36 86.44 113.7 159.0

\

@@

\
`          `\
\
\
`   `\
\
\
`   `

@@

\
`     `

    ##              Df Sum Sq Mean Sq F value Pr(>F)    
    ## series        3  39184   13061   61.17 <2e-16 ***
    ## Residuals   396  84553     214                   
    ## ---
    ## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

\
`     `

    ##              Df Sum Sq Mean Sq F value Pr(>F)    
    ## series        3  42229   14076   63.85 <2e-16 ***
    ## Residuals   396  87301     220                   
    ## ---
    ## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

\
\
`  `\
`       `

    ##                      W      p.value
    ## Treat.a - Treat.b  695 7.170082e-26
    ## Depth.1 - Depth.2 2064 7.360379e-13

\
`  `\
`       `

    ##                      W      p.value
    ## Treat.a - Treat.b  880 7.845518e-24
    ## Depth.1 - Depth.2 1599 9.672040e-17

>>>>

This demo aims to give quick overview of the <span>`dispRity`
</span> package (v.0.2) for ecological analysis. Please refer to [GitHub
page:
github.com/TGuillerme/dispRity](https://github.com/TGuillerme/dispRity)
for other vignettes, namely the `dispRity` tutorial that
explains the functions in more detail.

@@

```{r, eval = FALSE}
if(!require(devtools)) install.packages("devtools")
install_github("TGuillerme/dispRity", ref = "release")
library(dispRity)
```

@@

```{r}
## Loading demo and the package data
library(dispRity)

## Setting the random seed for repeatability
set.seed(123)

## Loading the data
data(McClean_data)

## This dataset contains an ordinated matrix (20 dimensions) of the distance
## between 40 experimental plots.
ord_matrix <- McClean_data$ordination

## As well as two list of different factors affecting each experimental plot:
## the treatment and the depth.
treatments <- McClean_data$treatment
depth <- McClean_data$depth
```

@@

```{r}
##<!--, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>= -->
## The x and y axis represent the two first dimensions (i.e. principal
## components) of the ordination
x <- ord_matrix[, 1]
y <- ord_matrix[, 2]
## The colors will represent the treatments
cols <- sub("a", "red", treatments)
cols <- sub("b", "blue", cols)
## The symbols will represent the depth
pchs <- sub(1, 16, depth)
pchs <- as.numeric(sub(2, 17, pchs))

## Graphical option
par(bty = "n")
## A classic 2D ordination plot
plot(x, y, col = cols, pch = pchs, xlab = "PC 1", ylab = "PC 2",
    xlim = range(x) + c(0, 100))
```

@@

```{r}
## Creating the table that contain the elements and their attributes
factors <- as.data.frame(matrix(data = c(treatments, depth), nrow = nrow(ord_matrix),
    ncol = 2, byrow = FALSE, dimnames = list(rownames(ord_matrix))))
names(factors) <- c("Treat", "Depth")
head(factors)
```

@@

```{r}
## Splitting the ordinated space into four subsamples
customised_series <- cust.series(ord_matrix, factors)

## Note that the output of dispRity functions are dispRity objects
class(customised_series)
## These objects are automatically printed in a summary method (calling S3 print.dispRity)
## giving information about the object
customised_series
```

@@

```{r}
## Summarising the object
str(customised_series)

## Displaying the full object
print(customised_series, all = TRUE)
```

@@

```{r}
## Calculating disparity
disparity <- dispRity(customised_series, metric = c(median, centroids))

## Note that disparity is a dispRity object and printing it just gives details
## on the object, not the results. We need to use summary.dispRity (S3) to get
## the results.
disparity
summary(disparity)
```

@@

```{r}
## Simple bootstrapping (100 times)
bootstrapped_data <- boot.matrix(customised_series, bootstraps = 100)

## Bootstrapping with rarefaction (i.e. only re-sampling 17 elements each time)
rarefied_data <- boot.matrix(customised_series, bootstraps = 100,
    rarefaction = 17)

## Note that the output is a dispRity object giving some details on the series
## and the bootstraps
bootstrapped_data
rarefied_data
```

@@

```{r}
## Bootstrapped disparity
disparity_BS <- dispRity(bootstrapped_data, metric = c(median, centroids))

## Rarefied disparity
disparity_rare <- dispRity(rarefied_data, metric = c(median, centroids))
```

@@

```{r}
## Summarising disparity (default)
summary(disparity_BS)

## The quantiles are calculated as 50 and 95 and the central tendency is the
## mean by default but we can specify different options
summary(disparity_BS, quantile = 90, cent.tend = median)

## Finally we can see the results of the rarefaction analysis:
summary(disparity_rare)

## Note that for the first three series, there is no observed disparity since
## the data was rarefied for these series.
```

@@

```{r}
## Graphical options
quartz(width = 10, height = 5) ; par(mfrow = (c(1,2)), bty = "n")

## Plotting the bootstrapped disparity
plot(disparity_BS, main = "Bootstrapped data")

## Plotting the rarefied disparity
plot(disparity_rare, main = "Rarefied data")
```

@@

```{r}
## Testing the effect of our factors on the bootstrapped data
summary(test.dispRity(disparity_BS, test = aov, comparisons = "all"))

## Testing the effect of our factors on the rarefied data
summary(test.dispRity(disparity_rare, test = aov, comparisons = "all"))

## Post-hoc testing of the effect of the two different factors
## Note that the comparison contains the list of the pairs of series to compare
test.dispRity(disparity_BS, test = wilcox.test,
    comparisons = list(c(1,2), c(3,4)))

## Testing the effect of the two different factors for the rarefied data
test.dispRity(disparity_rare, test = wilcox.test,
    comparisons = list(c(1,2), c(3,4)))
```
